---
title: "Disk-backed networks with parquetEdgeStore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Disk-backed networks with parquetEdgeStore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load, message = FALSE}
library(GiottoDisk)
```

`parquetEdgeStore` is the disk-backed counterpart to in-memory networks
(`nnNetObj@network`, `spatialNetworkObj@network`). It stores edges in
long-format parquet plus a small node-ID sidecar, supports lazy filter
pushdown via `[`, and materializes back to `igraph` / `data.table` /
`arrow` on demand. It's the storage layer behind
`createNetwork(..., output = "parquet")` and the auto-write path on
network setters for backed `giotto` objects.

## A small synthetic example

We use a 20-vertex undirected random graph throughout this article so
every chunk is runnable end-to-end. Substitute your own network /
giotto pipeline where the comments say so.

```{r ex-make}
set.seed(42)
ig <- igraph::sample_gnm(20, 50, directed = FALSE)
igraph::V(ig)$name <- paste0("cell_", sprintf("%02d", seq_len(20)))
igraph::E(ig)$weight <- runif(50)
ig
```

## Direct opt-in — `createNetwork(..., output = "parquet")`

The explicit way: set `output = "parquet"` on the `networkParam` and
`createNetwork()` returns a `parquetEdgeStore` directly. (We simulate
the call here with a `storeWrite` of the in-mem igraph, which is what
`.finalize_network` does under the hood.)

```{r quickstart-explicit}
# In real code:
#   p <- networkParam("sNN", k = 30, output = "parquet")
#   edge_store <- createNetwork(pca, param = p)
edge_store <- storeWrite(
    storeCreate(type = "parquetEdgeStore"),
    ig, type = "sNN"
)
edge_store
```

If a `gDirSource` is supplied, the artifact lands in the project vault
instead of a tempdir. With the auto-write described below, you usually
don't have to set this flag explicitly.

## Auto-write on backed gobjects

When a `giotto` object has a `gsource` backend attached, the network
setters (`setNearestNetwork()` / `setSpatialNetwork()`) automatically
write any in-memory igraph through to a `parquetEdgeStore` before
storing it. The user-facing call site is unchanged from the in-memory
case.

```{r autowrite, eval = requireNamespace("GiottoClass", quietly = TRUE)}
gdir <- file.path(tempdir(), "vignette_autowrite")
unlink(gdir, recursive = TRUE)

mat <- matrix(rpois(20 * 50, 2), nrow = 50, ncol = 20,
              dimnames = list(paste0("g_", 1:50),
                              paste0("cell_", sprintf("%02d", 1:20))))
g <- GiottoClass::createGiottoObject(expression = mat, backend = gdir)

nn <- methods::new("nnNetObj",
    network    = ig,           # in-memory igraph
    nn_type    = "sNN",
    name       = "sNN.demo",
    spat_unit  = "cell",
    feat_type  = "rna",
    provenance = "cell"
)
options("giotto.check_valid" = FALSE)
g <- GiottoClass::setNearestNetwork(g, nn, verbose = FALSE)

nn_back <- GiottoClass::getNearestNetwork(g, output = "nnNetObj",
    spat_unit = "cell", feat_type = "rna",
    nn_type   = "sNN", name = "sNN.demo")
class(nn_back@network)
nn_back@network
```

The auto-write fires when both:

1. `gobject@source` is non-null (a backend is attached), AND
2. The incoming `@network` slot is NOT already a `dataStore` (so we
   never double-write an already-disk-backed network).

For gobjects without a backend, the setter is a no-op on storage
shape: in-memory networks stay in memory.

## Bulk promotion — `sourceWrite(src, gobject)`

If you have an existing in-memory `giotto` (e.g. from an analysis that
predates the backed-pipeline switch), the whole thing can be promoted
in one shot:

```{r bulk-promote, eval = requireNamespace("GiottoClass", quietly = TRUE)}
# An in-memory gobject built the normal way
g_inmem <- GiottoClass::createGiottoObject(expression = mat)
g_inmem
class(g_inmem@source)   # NULL — no backend

# Promote to a backed project
gdir_bulk <- file.path(tempdir(), "vignette_bulk_promote")
unlink(gdir_bulk, recursive = TRUE)
g_backed <- sourceWrite(gDirSource(path = gdir_bulk), g_inmem)
class(g_backed@source)
class(GiottoClass::getExpression(g_backed)[])
```

Internally this builds a fresh `giotto` with the gsource attached as
the backend, then re-attaches every data subobject via
[`GiottoClass::setGiotto()`]. Each setter's existing backend-aware
write path handles the conversion: expression matrices land as
`parquetExprStore`, polygons as `parquetGeomStore`, points as
`parquetGeomStore`, NN and spatial networks as `parquetEdgeStore`.

The operation is idempotent at the subobject level — every backend-
aware setter checks `inherits(x@<slot>, "dataStore")` and skips re-
writing if already on disk. Calling `sourceWrite(src, g_backed)`
again with a fully-backed gobject doesn't grow the manifest or
re-write artifacts. Useful if some subobject snuck through in-memory
on a project that should be fully backed — re-running this call
catches the stray.

Subobjects whose setters don't yet have backend-aware write logic
(e.g. some image variants) stay in-memory; this is a known limitation
that doesn't affect correctness.

## Reading back

Three output modes.

```{r read-arrow}
# Lazy Arrow query (zero RAM, streaming downstream)
ds <- storeRead(edge_store, output = "arrow")
head(dplyr::collect(ds))
```

```{r read-tibble}
# Collected data.table with character node IDs (joined from sidecar)
head(storeRead(edge_store, output = "tibble"))
```

```{r read-igraph}
# In-memory igraph (vertex names preserved via V(g)$name)
g_back <- storeRead(edge_store, output = "igraph")
g_back
head(igraph::V(g_back)$name)
```

The `igraph` materialization ranks node ints down to a dense `1..V`
range over the subsetted edges before construction — so the in-memory
igraph only ever sees small int IDs, regardless of how large the
source universe is. `V(g)$name` carries the original character
barcodes.

## Subsetting

`[` follows igraph's adjacency semantics, with one-arg shorthand for
the common induced-subgraph case.

### Induced subgraph

```{r sub-induced}
v_set <- c("cell_01", "cell_02", "cell_05", "cell_10")
sub <- edge_store[v_set]
storeRead(sub, output = "tibble")
```

### Adjacency slice

```{r sub-slice}
storeRead(edge_store[c("cell_01"), c("cell_05", "cell_10")],
          output = "tibble")
```

### Endpoint filter

```{r sub-endpoint}
# Edges incident to cell_01 (undirected: matches either side)
storeRead(edge_store[, "cell_01"], output = "tibble")
```

### Named-args (same as positional)

```{r sub-named}
identical(
    storeRead(edge_store["cell_01", "cell_05"], output = "tibble"),
    storeRead(edge_store[from = "cell_01", to = "cell_05"],
              output = "tibble")
)
```

### Complement

```{r sub-negate}
# All edges NOT incident to cell_01
storeRead(edge_store[, "cell_01", negate = TRUE], output = "tibble")[1:5]
```

All `[` calls are lazy — they append a filter to the store's `@ops`
and return a new handle. No I/O happens until `storeRead()`.

### Undirected vs directed semantics

For `@directed = FALSE` stores (sNN, spatial, undirected Delaunay),
edges are stored once in canonical form (`from_id <= to_id`). The `[`
methods make this transparent — asymmetric filters automatically OR
both orientations:

```{r sub-undirected-symmetry}
# Same set of edges regardless of orientation:
nrow(storeRead(edge_store[c("cell_01"), c("cell_05")], output = "tibble"))
nrow(storeRead(edge_store[c("cell_05"), c("cell_01")], output = "tibble"))
```

For `@directed = TRUE` stores (kNN), slices stay strict — the
original orientation is preserved.

```{r sub-directed}
ig_dir <- igraph::make_graph(c("a","b","b","c","c","a","d","a"),
                             directed = TRUE)
store_dir <- storeWrite(
    storeCreate(type = "parquetEdgeStore"),
    ig_dir, type = "kNN"
)
storeRead(store_dir["a", "b"], output = "tibble")
storeRead(store_dir["b", "a"], output = "tibble")  # empty: no b->a
```

## Node-ID sidecar

The sidecar is a small parquet at `<store_root>/nodes/` that maps
character node IDs (cell barcodes, point IDs) to dense integer IDs
used inside the edge file. It's auto-generated at write time.

```{r sidecar}
storeRead(edge_store@nodes, output = "tibble") |> head()
```

You don't normally interact with it — `[` and `storeRead` handle the
char ↔ int translation transparently. It exists so that the
in-memory class doesn't carry a billion-element character vector for
the vertex universe; instead the universe lives on disk and is joined
when needed.

### Writer-supplied node universe

When constructing the store directly via `storeWrite()`, an optional
`node_universe` argument preserves isolated vertices:

```{r node-universe}
# Edges only mention 3 vertices; full universe has 5
small_dt <- data.table::data.table(
    from = c("v1", "v1"), to = c("v2", "v3")
)
s <- storeWrite(
    storeCreate(type = "parquetEdgeStore"),
    small_dt,
    type = "spatial",
    directed = FALSE,
    node_universe = c("v1", "v2", "v3", "v4", "v5")
)
s@n_cells   # 5 — including isolated v4, v5
storeRead(s@nodes, output = "tibble")
```

Without `node_universe`, the sidecar derives from `unique(c(from, to))`
on the edge table — vertices with no edges would not appear.

## Tile-and-loop iteration

For per-tile spatial work (Moran's I, neighborhood enrichment), the
right shape is to load one tile's worth of edges into an in-memory
graph, then run the inner algorithm in a loop within that tile:

```r
# pseudocode — uses a separately built tile plan
plan <- tilework::quadtreePlan(geom_store, threshold = 50000)

results <- lapply(seq_along(plan), function(i) {
    v_set  <- storeRead(geom_store[plan[i]], output = "tibble")$cell_id
    g_tile <- storeRead(edge_store[v_set], output = "igraph")
    # ... inner per-cell loop on g_tile ...
})
```

One materialization per tile, amortized over many inner ops. The rank
step inside `storeRead(output = "igraph")` is cheap at typical chunk
sizes (~30 ms at 100k edges) and invisible at the algorithm level.

## Direct construction (escape hatches)

When the network is already in memory and you want to disk-back it
without going through `createNetwork()`:

```{r escape-igraph}
# From an in-memory igraph (no marker class needed)
store_from_igraph <- storeWrite(
    storeCreate(type = "parquetEdgeStore"),
    ig, type = "sNN"
)
store_from_igraph
```

```{r escape-edgedt}
# From an arbitrary edge data.table with non-standard col names
my_dt <- data.table::data.table(
    src = c("a", "a", "b"),
    dst = c("b", "c", "c"),
    w   = c(0.5, 0.7, 0.9)
)
inp <- edgeDTInput(my_dt, from_col = "src", to_col = "dst", weight_col = "w")
storeWrite(storeCreate(type = "parquetEdgeStore"), inp,
           type = "kNN", directed = TRUE)
```

These input markers are escape hatches — the primary path for
production work is `createNetwork(..., output = "parquet")` (or the
backed-gobject auto-write). They exist for migration from existing
pipelines, round-trip tests, and cases where the network was built
outside `createNetwork`.

## Snapshot integration

`parquetEdgeStore` participates in `gDirSource` snapshot lifecycle the
same as any other disk-backed store. With the auto-write on setters,
the typical flow is:

```r
g <- createGiottoObject(expression = mat, backend = "~/projects/my_atlas")

# build + slot network — auto-writes through to vault
nn <- createNetwork(g, param = networkParam("sNN", k = 30))
g <- setNearestNetwork(g, nn)

# snapshot ties the artifact to the giotto save
snapshotSave(g@source, g)

# later session: artifact is recovered from the manifest, no rebuild
g <- snapshotLoad("~/projects/my_atlas")
```

The store directory contains `edges/` and `nodes/` subdirectories;
future hive-partitioned writes (e.g. `edges/source_id=<uid>/...`)
drop in without changing the read API.

## When to use `parquetEdgeStore` vs in-memory igraph

| Scenario | Recommendation |
|---|---|
| < 10k cells, single session | in-memory igraph |
| 10k–500k cells, fits in RAM but heavy | either; parquet saves RAM during multi-step pipelines |
| 500k–10M cells | parquet — in-memory becomes a real bottleneck |
| 10M+ cells | parquet, tiled construction |
| Pipelines that round-trip to disk between R sessions | parquet — survives `snapshotSave` |

For Leiden clustering and other whole-graph compute, you can still
`storeRead(edge_store, output = "igraph")` to get an in-memory graph
when needed — the disk-backed store is the persistent canonical form,
not a replacement for in-memory graph compute.

## See also

- [`createNetwork()`](https://drieslab.github.io/GiottoClass/) on
  GiottoClass for the in-memory and parquet-output dispatch
- `parquetExprStore` — disk-backed expression matrices, same pattern
- `parquetGeomStore` — disk-backed spatial points / polygons
