---
title: "Disk-backed networks with parquetEdgeStore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Disk-backed networks with parquetEdgeStore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`parquetEdgeStore` is the disk-backed counterpart to in-memory networks
(`nnNetObj@network`, `spatialNetworkObj@network`). It stores edges in
long-format parquet plus a small node-ID sidecar, supports lazy filter
pushdown via `[`, and materializes back to `igraph` / `data.table` /
`arrow` on demand. It is the storage layer that `createNetwork()` writes
into when called with `output = "parquet"`.

This article walks through the common patterns. For the architectural
rationale see the proposal at
`scstream_prototype/proposals/parquetEdgeStore.md`.

## Quick start — via `createNetwork()`

The primary entry point is `Giotto::createNetwork()` with a
`networkParam` whose `output` slot is `"parquet"`:

```{r}
library(Giotto)
library(GiottoDisk)

# kNN over PCA scores, output as a disk-backed network store
p <- networkParam("kNN", k = 30, output = "parquet")
edge_store <- createNetwork(my_pca_obj, param = p)

edge_store
#> <parquetEdgeStore> type=kNN directed=TRUE
#>   n_cells: 76,952  n_edges: 2,308,560
#>   path:    /tmp/.../gdisk_dump/file...
```

If your project lives under a `gDirSource`, point `createNetwork()` at
the backend so the artifacts land in the vault instead of a tempdir:

```{r}
src <- gDirSource(path = "~/projects/my_atlas")
edge_store <- createNetwork(my_pca_obj, param = p, backend = src)
```

The resulting `parquetEdgeStore` slots directly into a `nnNetObj`
without conversion — `@network` is type `ANY`:

```{r}
nn <- new("nnNetObj", network = edge_store, nn_type = "kNN")
g <- setGiotto(g, nn)
```

## Reading back

Three output modes:

```{r}
# Lazy Arrow query (zero RAM, streaming downstream)
ds <- storeRead(edge_store, output = "arrow")

# Collected data.table with character node IDs (joined from sidecar)
dt <- storeRead(edge_store, output = "tibble")

# In-memory igraph (vertex names preserved via V(g)$name)
g <- storeRead(edge_store, output = "igraph")
```

The `igraph` materialization ranks node ints down to a dense `1..V` range
over the subsetted edges before construction — so the in-memory igraph
only ever sees small int IDs, regardless of how large the source
universe is. `V(g)$name` carries the original character barcodes.

## Subsetting

`[` follows igraph's adjacency semantics, with one-arg shorthand for
the common induced-subgraph case:

```{r}
# Induced subgraph on a vertex set (both endpoints in the set)
sub <- edge_store[c("cell_A", "cell_B", "cell_C")]

# Adjacency slice — edges from one set to another
slice <- edge_store[from_cells, to_cells]

# Endpoint filter
incoming <- edge_store[, "cell_X"]      # edges ending at cell_X
outgoing <- edge_store[from = "cell_X"] # edges starting at cell_X

# Explicit named form
slice <- edge_store[from = from_cells, to = to_cells]

# Negate any of the above
not_this <- edge_store[from = "cell_X", negate = TRUE]
```

All `[` calls are lazy — they append a filter to the store's `@ops` and
return a new handle. No I/O happens until `storeRead()`. You can chain:

```{r}
storeRead(edge_store[v_set][weight > 0.5], output = "igraph")
```

### Undirected vs directed semantics

For `@directed = FALSE` stores (sNN, spatial, undirected Delaunay), edges
are stored once in canonical form (`from_id <= to_id`). The `[` methods
make this transparent — asymmetric filters automatically OR both
orientations:

```{r}
# Undirected store: orientation-agnostic
g <- networkParam("sNN", k = 30, output = "parquet") |> createNetwork(pca)
g[, "cell_X"]        # matches edges where cell_X is either endpoint
g[v_a, v_b]          # matches both v_a -> v_b and v_b -> v_a edges
```

For `@directed = TRUE` stores (kNN), slices stay strict — the original
orientation is preserved.

```{r}
g <- networkParam("kNN", k = 30, output = "parquet") |> createNetwork(pca)
g[, "cell_X"]        # only edges directed TO cell_X
g["cell_A", "cell_B"] # only the cell_A -> cell_B edge
```

## Tile-and-loop iteration

For per-tile spatial work (Moran's I, neighborhood enrichment), the
right shape is to load one tile's worth of edges into an in-memory
graph, then run the inner algorithm in a loop within that tile:

```{r}
plan <- tilework::quadtreePlan(geom_store, threshold = 50000)

results <- lapply(seq_along(plan), function(i) {
    v_set <- storeRead(geom_store[plan[i]], output = "tibble")$cell_id
    g_tile <- storeRead(edge_store[v_set], output = "igraph")
    # ... inner per-cell loop on g_tile ...
})
```

One materialization per tile, amortized over many inner ops. The rank
step inside `storeRead(output = "igraph")` is cheap at typical chunk
sizes (~30 ms at 100k edges) and invisible at the algorithm level.

## Node-ID sidecar

The sidecar is a small parquet at `<store_root>/nodes/` that maps
character node IDs (cell barcodes, point IDs) to dense integer IDs used
inside the edge file. It's auto-generated at write time.

```{r}
storeRead(edge_store@nodes, output = "tibble")
#>    row_index    node_id int_id
#>        <int>     <char>  <int>
#> 1:         1     cell_1      1
#> 2:         2     cell_2      2
#> ...
```

You don't normally interact with it — `[` and `storeRead` handle the
char ↔ int translation transparently. It exists so that the in-memory
class doesn't carry a billion-element character vector for the vertex
universe; instead the universe lives on disk and is joined when needed.

### Writer-supplied node metadata

When constructing the store directly via `storeWrite()`, an optional
`node_universe` argument preserves isolated vertices:

```{r}
all_cells <- pDataDT(g)$cell_ID   # full universe, including isolates
storeWrite(
    storeCreate(type = "parquetEdgeStore"),
    my_edge_dt,
    type = "spatial",
    directed = FALSE,
    node_universe = all_cells
)
```

Without `node_universe`, the sidecar derives from `unique(c(from, to))`
on the edge table — vertices with no edges would not appear.

## Direct construction (escape hatches)

When the network is already in memory and you want to disk-back it
without going through `createNetwork()`:

```{r}
# From an existing data.table of edges
inp <- edgeDTInput(
    my_edge_dt,
    from_col = "source", to_col = "target",
    weight_col = "weight"
)
edge_store <- sourceWrite(src, inp, store_type = "parquetEdgeStore")

# From an existing in-memory igraph (round-trip / migration)
inp <- igraphInput(g, weight_attr = "weight")
edge_store <- sourceWrite(src, inp, store_type = "parquetEdgeStore")

# From an RANN / dbscan kNN result
nn <- dbscan::kNN(pca_scores, k = 30)
inp <- nnSearchInput(nn, cell_ids = rownames(pca_scores))
edge_store <- sourceWrite(src, inp, store_type = "parquetEdgeStore")
```

These input markers are escape hatches — the primary path is
`createNetwork(..., output = "parquet")`. They exist for migration from
existing pipelines, round-trip tests, and cases where the network was
built outside Giotto's `createNetwork` machinery.

## Snapshot integration

`parquetEdgeStore` participates in `gDirSource` snapshot lifecycle the
same as any other disk-backed store:

```{r}
src <- gDirSource(path = "~/projects/my_atlas")
edge_store <- createNetwork(pca, param = p, backend = src)

# saves the edge_store's artifact alongside the giotto snapshot
snapshotSave(src, my_gobject)

# later session
g <- snapshotLoad(src)
# edge_store handles are restored from manifest; no rebuild
```

The store directory contains `edges/` and `nodes/` subdirectories;
future hive-partitioned writes (e.g. `edges/source_id=<uid>/...`) drop
in without changing the read API.

## When to use `parquetEdgeStore` vs in-memory igraph

| Scenario | Recommendation |
|---|---|
| < 10k cells, single session | in-memory igraph |
| 10k–500k cells, fits in RAM but heavy | either; parquet saves RAM during multi-step pipelines |
| 500k–10M cells | parquet — in-memory becomes a real bottleneck |
| 10M+ cells | parquet, tiled construction |
| Pipelines that round-trip to disk between R sessions | parquet — survives `snapshotSave` |

For Leiden clustering and other whole-graph compute, you can still
`storeRead(edge_store, output = "igraph")` to get an in-memory graph
when needed — the disk-backed store is the persistent canonical form,
not a replacement for in-memory graph compute.

## See also

- [`createNetwork()`](https://drieslab.github.io/GiottoClass/) on
  GiottoClass for the in-memory and parquet-output dispatch
- `parquetExprStore` — disk-backed expression matrices, same pattern
- `parquetGeomStore` — disk-backed spatial points / polygons
- The architectural rationale at
  `scstream_prototype/proposals/parquetEdgeStore.md`
